---
title: "Assignment 5: Data Visualization"
author: "Elizabeth Keeley"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

## OVERVIEW

This assignment accompanies the Environmental Data Analytics module on
data visualization. You will create, format, combine, and interpret
plots using environmental datasets. Your goal is not only to produce
correct code, but also to make deliberate choices about themes, color,
accessibility, and the message communicated by each figure.

## Directions

1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing
    `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill
    each instruction.
4.  Be sure your code is tidy; use line breaks to ensure your code fits
    in the knitted output.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code
    into a single PDF file.
7.  All figures should include clear labels and units, readable text,
    and color choices appropriate for the type of variable displayed. Do
    not rely on color alone when another cue (such as shape, linetype,
    or facets) would make the figure more accessible.
8.  Complete your own code before using AI in the designated comparison
    activity in Question 7. Include the prompt and your evaluation of
    the AI response.

------------------------------------------------------------------------

## Set up your session

1.  Set up your session. Load the `tidyverse`, `here`, `cowplot`, and
    any color palette package(s) you choose to use. Verify your project
    directory. Read in the processed NTL-LTER nutrients and
    chemistry/physics file for Peter and Paul Lakes
    (`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`) and
    the processed Niwot Ridge litter file
    (`NEON_NIWO_Litter_mass_trap_Processed.csv`) from the
    `Processed_KEY` folder.

2.  Make sure R is reading dates as date format; if not change the
    format to date.

```{r initialize project, message=FALSE, warning=FALSE}
#1 
library(tidyverse)
library(here)
library(cowplot)

# read in processed data
peterpaul <- read.csv(
  here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")) 

litter <- read.csv(
  here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"))

#2 correct dates to date format
peterpaul$sampledate <- ymd(peterpaul$sampledate)
litter$collectDate <- ymd(litter$collectDate)

```

## Define your theme

3.  Build a theme and set it as your default theme. Start with a
    complete theme, such as `theme_minimal()` or `theme_classic()`, and
    then customize at least two of the following:

-   Plot background
-   Plot title
-   Axis labels
-   Axis ticks/gridlines
-   Legend

Choose a base font size that will remain readable when the plots are
knitted. Briefly identify the two or more elements you changed and
explain how the changes improve readability.

```{r build plot theme, message=FALSE, warning=FALSE}
#3 

keeley_theme <- theme_minimal(base_size = 12)  + 
  theme(
    # Modified inheritance structure of text element
    plot.title =        element_text(#size = 14,
                                     face = "bold", 
                                     hjust = 0.5),
    plot.subtitle = element_text(#size = 13, 
                                 hjust = 0.5),
    axis.title.x =      element_text(size = 12),
    axis.title.y =      element_text(size = 12),
    axis.text =         element_text(size = 11),
    
    # Modified inheritance structure of rect element
    panel.background =  element_rect(color = "grey"),
    legend.background =        element_rect(color = "grey"),
    
    
    # Modifying legend.position
    legend.position = "bottom"
  )

```

> Theme choices and rationale: I centered the plot title and subtitle because I find when the titles are left-aligned, that the words can get lost next to the y-axis labels. I adjusted the text sizes of the x and y axes and the axis labels because I find that having them different sizes is useful for separating the text between axis title and axis label. I put the legend at the bottom of the plot because it allows more room for the plot to fill the horizontal space, which I think makes it more legible. 

## Create graphs

For Questions 4–7, create `ggplot2` graphs and adjust their aesthetics
using data-visualization best practices. Match the color palette to the
variable type, label variables and units clearly, and remove formatting
that does not support the comparison.

4.  [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`),
    using separate aesthetics for Peter and Paul lakes. Do not rely only
    on color: also use shape, linetype, facets, or another appropriate
    visual cue. Add a separate line of best fit for each lake using the
    `lm` method. Zoom in so that the central pattern is visible without
    deleting the extreme observations from the analysis (hint: use
    `coord_cartesian()` rather than `xlim()` or `ylim()`). Include a
    descriptive title, axis labels with units, and a meaningful legend
    title.

```{r plot total P vs PO4, message=FALSE, warning=FALSE}
#4 Setting up the total phosphorus and phosphate plot. 

PvP04 <- peterpaul %>% 
  ggplot(aes(x=po4, y=tp_ug, color = lakename))+
  geom_point(size=1, 
             alpha = 0.5)+
  scale_color_viridis_d(option = "plasma", end=0.5)+
  geom_smooth(method="lm", 
              se=FALSE, linewidth = 0.3)+
  facet_wrap(vars(lakename), nrow=2)+
  labs(title = "Relationship between Total Phosphorous Concentration
       and Phosphate Concentration",
       subtitle = "Peter and Paul Lakes",
       y= "Total Phosphorus Concentration (µg P/L)",
       x= "Phospate Concentration (µg P/L)",
       color = "Lake Name")+
  coord_cartesian(xlim = c(0, 45),#150),
                 ylim = c(0,150))+
  keeley_theme
  
PvP04      

```

Question: What relationship do you observe between phosphate and total
phosphorus? Does the relationship appear to differ between lakes?
Identify any extreme observations and explain why zooming with
`coord_cartesian()` is preferable to removing them with scale limits.

> Answer: There is a positive relationship between phosphate and total
> phosphorus in both lakes. The relationship, if using the line of best
> fit, does appear to differ between lakes. When looking at the trend of
> observations between 0 and 50 micrograms, the relationship does not
> appear hugely different. However, there is one observation in Paul
> Lake where the phosphate concentration is over 300 micrograms. This
> observation substantially impacts the line of best fit for the Paul
> Lake relationship. Zooming in with coord_cartesian() allows us to
> examine the trends in a way that is more visually accessible without
> excluding more extreme observations from the trendline analysis. We
> can decide to exclude that observation later, but it is useful to keep
> it until we decide the observation is an outlier that can be excluded.

5.  [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP,
    and (c) TN, with month on the x-axis and lake represented by a
    consistent aesthetic. Use a qualitative, color-vision-friendly
    palette. Then create a `cowplot` that combines the three graphs.
    Make sure that only one legend is present, the month order is
    correct, graph axes are aligned, and each y-axis states the variable
    and units.

Tips: \* Month should appear in calendar, not alphabetical, order. You
will need to convert to factor to order it. \* Setting an axis title in
your theme to `element_blank()` removes the axis title (useful when
multiple, aligned plots use the same axis values) \* Setting a legend's
position to "none" will remove the legend from a plot. \* Individual
plots can have different sizes when combined using `cowplot`.

```{r Create boxplots, fig.width=10, fig.height=11, message=FALSE, warning=FALSE}
#5 
# make boxplot for temp

temp <- peterpaul %>% 
  mutate(month = month(sampledate)) %>% 
  mutate(month = factor(month)) %>% 
  filter(month != 2) %>% 
  ggplot(aes(x = month, 
             y = temperature_C, 
             color = lakename, 
             pattern=lakename))+
  geom_boxplot(outlier.size=0.6, 
               alpha = 0.5, 
               linewidth=0.5,
               fatten=0.6) +
  scale_color_viridis_d(option = "plasma", 
                        end=0.5)+
  labs(title = "Temperature, Total Phosphorus, and Total Nitrogen",
       subtitle= "Medians and Quartiles by Month and Lake",
       y = "Temperature\n (C)")+
  keeley_theme+
  theme(axis.title.x = element_blank(),
        legend.position = "none")

# make boxplot for total phosophorus concentration
TP <-  peterpaul %>% 
  mutate(month = month(sampledate)) %>% 
  mutate(month = factor(month)) %>%
  filter(month != 2) %>% 
  ggplot(aes(x = month, 
             y = tp_ug, 
             color = lakename)) +
  geom_boxplot(size =0.5, 
               outlier.size=0.6, 
               alpha = 0.5, 
               linewidth=0.5, 
               fatten=0.6) +
  scale_color_viridis_d(option = "plasma", 
                        end=0.5)+
  labs(y = "Total Phosphorus\n Concentration\n (µg P/L)")+
  coord_cartesian(ylim = c(0,125))+
  keeley_theme+
  theme(axis.title.x = element_blank(),
        legend.position = "none")

# make boxplot for total nitrogen concentration
TN <-  peterpaul %>% 
  mutate(month = month(sampledate)) %>% 
  mutate(month = factor(month)) %>% 
  filter(month != 2) %>% 
  ggplot(aes(x = month, 
             y = tn_ug, 
             color = lakename)) +
  geom_boxplot(outlier.size=0.6, 
               alpha = 0.5, 
               linewidth=0.5,
               fatten=0.6) +
  scale_color_viridis_d(option = "plasma", end=0.5)+
  labs(y= "Total Nitrogen\n Concentration\n (µg P/L)", x = "Month")+
  coord_cartesian(ylim = c(0,2500))+
  keeley_theme+
  theme(legend.title = element_blank())

# use cowplot to facet these figures
tempTNTP <- plot_grid(temp, TP, TN, 
                      ncol = 1, align = 'v', 
                      rel_heights = c(1, 0.8, 1.3))
tempTNTP

```

Question: What do you observe about the center, spread, and potential
outliers of each variable across seasons and between lakes? Support your
interpretation with specific features visible in the boxplots.

> Answer: The center and spread for temperature seems to be generally
> similar between the two lakes, with the exception of October.
> Temperature center increases from May to September, and then decreases
> from September to November. For total phosphorus, there seem to be a
> lot more outliers than for temperature, and those outliers are spread
> across a greater range of concentrations for Peter Lake than for Paul
> Lake. This is visible from the outlier points present on the center
> plot. The center is consistently higher in Peter Lake than in Paul
> Lake, and this does not seem to vary much by season. The spread is
> generally larger for Peter Lake than for Paul Lake, except for in May
> where the spread looks to be about the same between the two lakes. The
> outliers for Paul Lake have a greater range of concentration June
> through July than they do in May. There are no observations present
> for October and November. For total nitrogen, the outliers for Peter
> Lake also have a greater range in concentration than they do in Paul
> Lake. Nitrogen concentration is higher than Phosphorus concentration
> across season and lake. The center is similar between Paul and Peter
> Lakes for all months where there are observations except for August,
> where the center is higher for Peter Lake.

6.  [Niwot Ridge] Subset the litter dataset to the `"Needles"`
    functional group. Plot dry mass of needle litter by date and
    distinguish NLCD classes using a qualitative color aesthetic. You do
    not need to rename the land-use classes. Format the date axis,
    labels, legend, point size, and transparency so that overlapping
    observations remain visible.

7.  [Niwot Ridge] First, independently create a second version of the
    plot with the three NLCD classes separated into facets rather than
    distinguished by color. Keep other formatting choices as consistent
    as possible so that you can compare the effect of faceting itself.

```{r Plot litter, message=FALSE, warning=FALSE}
#6 Distinguishing nlcd class by color

drymass <- litter %>% 
  filter(functionalGroup == "Needles") %>% 
  ggplot(aes(x=collectDate, y=dryMass, color=nlcdClass))+
  geom_point(size = 1.5,alpha=0.5)+
  scale_color_viridis_d("NLCD Class")+
  labs(title = "Needle Litter Dry Mass by Date and NLCD Class",
       x = "Collection Date", 
       y= "Dry Mass (g)")+
  keeley_theme+
  theme(legend.position = "bottom")

drymass

#7 Distinguishing nlcd class by facet

drymass_faceted <- litter %>% 
  filter(functionalGroup == "Needles") %>% 
  ggplot(aes(x=collectDate, y=dryMass))+
  geom_point(size = 1.5,alpha=0.5)+
  facet_wrap(vars(nlcdClass), nrow=3) +
  #scale_color_viridis_d("NLCD Class")+
  labs(title = "Needle Litter Dry Mass by Date and NLCD Class",
       x = "Collection Date", 
       y= "Dry Mass (g)")+
  keeley_theme+
  theme(legend.position = "bottom")

drymass_faceted

```

## AI comparison

8.  After completing your own faceted plot, provide an AI tool with the
    Question 7 instructions and the relevant variable names. Ask it to
    propose `ggplot2` code for the same figure. Do **not** replace your
    original solution with the AI version.

Paste the exact prompt you used below:

> AI tool used: Google AI

> AI prompt: Suggest code for the following prompt: "Plot dry mass of
> needle litter by date with the three NLCD classes separated into
> facets. You do not need to rename the land-use classes. Format the
> date axis, labels, legend, point size, and transparency so that
> overlapping observations remain visible." The data frame is called
> litter. There are multiple functional groups. Please select for Needle
> litter only. functional groups: functionalGroup NLCD classes :
> nlcdClass\
> dry mass : dryMass\
> date : collectDate

Run and inspect the AI-generated code, correcting only errors required
to make it run. If you make corrections, identify them. Then compare the
two solutions.

```{r AI solution, message=FALSE, warning=FALSE}
# Paste the AI-generated code here. Clearly comment any corrections you make.

library(ggplot2)
library(scales)

# Filter for Needle litter and create the plot
litter_filtered <- subset(litter, functionalGroup == "Needles") 
#the AI had functionalGroup as "Needle Litter" so I changed that to "Needles"

ggplot(litter_filtered, aes(x = as.Date(collectDate), y = dryMass)) +
  # Alpha handles transparency; size makes points manageable for overlapping data
  geom_point(alpha = 0.5, size = 1.5, color = "darkgreen") +
  # Separate into facets based on NLCD classes
  facet_wrap(~ nlcdClass, ncol = 3) +
  # Format the date axis cleanly
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  # Add clear labels and formatting
  labs(
    title = "Dry Mass of Needle Litter over Time by NLCD Class",
    x = "Collection Date",
    y = "Dry Mass (g)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotates date labels to prevent overlap
    panel.spacing = unit(1, "lines"),
    strip.text = element_text(face = "bold", size = 11)
  )

```

9.  Which plot—your solution or the AI solution—more effectively
    communicates the data? Evaluate at least three specific choices,
    such as data filtering, facet structure, date-axis formatting,
    labels and units, theme, accessibility, or unnecessary code.
    Identify one useful idea from the AI solution and one choice you
    would reject or improve. If the solutions are very similar, explain
    why both approaches are appropriate.

> AI comparison: I prefer the AI solution's choice to facet the three
> plots in one row. It makes the comparison between sites of Dry Mass
> much easier to quickly see. I think the collection dates along the x
> axis in the AI solution are too busy and hard to read, but I like that
> it provides more information about the dates than my solution does. I
> also like that the AI solution includes color because it makes the
> plot feel more interesting to look at. I would take the faceting that
> the AI chose, and I would reject the x-axis labeling and find a happy
> medium between my too-few labels the the AI solution's too-many
> labels.

10. Question: Which encoding in Questions 6 and 7—color or facets—do you
    think is more effective for comparing NLCD classes, and why?
    Consider overlap, ease of comparison, accessibility, and the amount
    of space used by each figure.

> Answer: I think the facets would be more helpful if I had faceted with
> one row instead of one column. It's a little hard to note trends or
> major differences or similarities between the three NLCD classes with
> all the observation points intermixed, so I prefer the faceted version
> for quick ease of interpretation. I would also choose to both facet
> and keep the colors different for each NLCD class.
