---
title: "Assignment 3: Data Exploration"
author: "Elizabeth Keeley"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Exploration M3: Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2016-26_raw.csv). Name these datasets "Neonics" and "Litter", respectively.

```{r setup,  message = FALSE, warning = FALSE}

#Import libraries
library(tidyverse)
library(lubridate)
library(knitr)
library(here)

#Check current working directory
here()

# Upload ECOTOX neonicotinoid dataset
Neonics <- read.csv(here("Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.txt"), sep = '|')

# Upload Niwot Ridge NEON dataset for litter and woody debris
Litter <- read.csv(here("Data/Raw/NEON_NIWO_Litter_massdata_2016-26_raw.csv"))
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Given that neonicotinoids are used widely in agriculture, it is important to udnerstand their impact on the insects they are intended to target as well as their unintended impact on other insects. It is likely that insects not detrimental to agriculture are harmed and killed by neonicotinoids, and so it is important to understand the broader implications of their use.  

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Litter and woody debris in forests is interesting for multiple reasons. Quantifying litter and woody debris, and the rate at which they decompose, can tell us something about carbon storage vs carbon release in those ecosystems. This data set appears to focus on dry mass, approximately 50% of which is typically carbon biomass. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Ground traps are used to collect fine woody debris whereas elevated traps are used to collect litter.
 2. Litter is material dropped from the forest canopy with an end diameter of less than 2cm and a length of less than 50cm. 
 3. Fine woody debris is material with an end diameter of less than 2cm but a length of greater than 50cm. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Q5}
# dimensions of Neonics
dim(Neonics) # this dataset has 8748 rows and 76 columns 

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Q6}
# Use summary and sort to determine the most common effects studied
summary(factor(Neonics$Effect))
sort(summary(factor(Neonics$Effect)), decreasing = TRUE) 
```

> Answer: The most common effects are Mortality, Population, Genetics, Behavior, and Feeding Behavior (in descending order). Mortality is of interest because it communicates both the effectiveness of the insecticide on target insecys, and the amount of death caused to other insect populations as a result. Population is 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r Q7}
# Use sort and summary to determine the most commonly studied species in the data set
sort(summary(factor(Neonics$X.Species.Common.Name), maxsum= Inf), decreasing = TRUE)

```

> Answer: The six most commonly studied species in this dataset are Honey Bee, Carniolan Honey Bee, Parasitic Wasp, Bumble Bee, Italian Honeybee, and Asiatic Honey Bee. These species are all pollinators, and thus the impact of insecticides on them is important to understand given widespread concerns over a  decline in pollinator species impacting agriculture and other plant survival. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1.Mean..Author..` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Q8}
# View Neonics data frame
view(Neonics)
# determine class for Conc.1.Mean..Author..
class(Neonics$Conc.1.Type..Author..) # the class is character 

```

> Answer: The class for this data column is character. 

## Explore your data graphically (Neonics)

9. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Q11}


```

> Answer:

## Explore your data (Litter)

10. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Q12}


```

11. Using the `unique` function, list the different plotIDs sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Q13}


```

> Answer:

12. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Q14}


```


What type(s) of litter tend to have the highest biomass at these sites?

> Answer:

13. Write a short data-exploration summary for Neonics.

> - **One key characteristic:** 
> - **One data-quality issue:** 
> - **One interesting pattern:** 
> - **One question for further analysis:** 

14. Write a short data-exploration summary for Litter.

> - **One key characteristic:** 
> - **One data-quality issue:** 
> - **One interesting pattern:** 
> - **One question for further analysis:** 
