---
title: "Assignment 6: Data Scraping"
author: "Elizabeth Keeley"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 Loading packages and checking working directory
library(tidyverse)
library(rvest)
library(here)
library(lubridate)
library(dplyr)
here()

#setting theme for plots
keeley_theme <- theme_minimal(base_size = 12)  + 
  theme(
    # Modified inheritance structure of text element
    plot.title =        element_text(#size = 14,
                                     face = "bold", 
                                     hjust = 0.5),
    plot.subtitle = element_text(#size = 13, 
                                 hjust = 0.5),
    axis.title.x =      element_text(size = 12),
    axis.title.y =      element_text(size = 12),
    axis.text =         element_text(size = 11),
    
    # Modified inheritance structure of rect element
    panel.background =  element_rect(color = "grey"),
    legend.background =        element_rect(color = "grey"),
    
    # Modifying legend.position
    legend.position = "bottom"
  )

theme_set(keeley_theme)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2025
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
webpage <- 
  read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2025") 

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 Assign values to be scraped to variables

system_name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()

pwsid <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()

ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()

max_day_use <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 Create a data frame from scraoed data

df_wateruse <- data.frame(
  "Month" = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12),
  "Year" = 2025,
  "System_Name" = system_name,
  "Ownership" = ownership, 
  "PWSID" = pwsid,
  "Maximum_Day_Use" = as.numeric(max_day_use)) %>% 
    mutate("Date" = my(paste(Month,"/",Year))
)

#5  Plot the data in the above dataframe

wateruse <- df_wateruse %>% 
  ggplot(aes(x=(Month), y=Maximum_Day_Use, group=1))+
  geom_line()+
  scale_x_continuous(breaks=1:12,
                    labels=month.abb)+
  labs(title = "Maximum Daily Withdrawal of Water",
       subtitle= "Durham, 2025",
       x = "Month", 
       y = "Maximum Daily Use (MGD)")

wateruse

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. Create a scraping function

scrape_func <- function(the_pwsid, the_year){
  #Get the proper url
  the_url <- 
    read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?', 
                                  'pwsid=',the_pwsid,"&year=",the_year))
  
  # scrape data 
  system_name <- the_url %>% 
    html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
    html_text()
  
  ownership <- the_url %>% 
    html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
    html_text()
  
  max_day_use <- the_url %>% 
    html_nodes("th~ td+ td") %>% 
    html_text()
  
  # create data frame
  df_wateruse2 <- data.frame(
  "Month" = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12),
  "Year" = rep(the_year,12), 
  "System_Name" = system_name,
  "Ownership" = ownership, 
  "PWSID" = the_pwsid,
  "Maximum_Day_Use" = as.numeric(max_day_use)) %>% 
  mutate("Date" = my(paste(Month,"/",Year)))
  
  #return data frame
  return(df_wateruse2)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 Use scraping function to extract and then plot withdrawals

durham_max_withd <- scrape_func("03-32-010", 2020) 

plot_dur_withd <- durham_max_withd %>% 
  ggplot(aes(x=(Month), y=Maximum_Day_Use, group=1))+
  geom_line()+
  scale_x_continuous(breaks=1:12,
                    labels=month.abb)+
  labs(title = "Maximum Daily Withdrawal of Water",
       subtitle= paste("Durham,",durham_max_withd$Year),
       x = "Month", 
       y = "Maximum Daily Use (MGD)")

plot_dur_withd

glimpse(durham_max_withd)

```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 Use scraping function to extract and plot Cary and Durham data from 2020

cary_maxwithd <- scrape_func('03-92-020', 2020)

max_withd <- bind_rows(cary_maxwithd, 
                       durham_max_withd)

plot_withd_comp <- max_withd %>% 
  ggplot(aes(x=(Month), 
             y=Maximum_Day_Use, 
             color = System_Name, 
             linetype= System_Name))+
  geom_line(linewidth = 0.7)+
  scale_x_continuous(breaks=1:12,
                    labels=month.abb)+
  labs(title = "Maximum Daily Withdrawal of Water",
       subtitle= paste("Durham and Cary,",max_withd$Year),
       x = "Month", 
       y = "Maximum Daily Use (MGD)",
       color = "System Name",
       linetype = "System Name")+
   scale_color_viridis_d(option = "plasma", 
                        end=0.5)

plot_withd_comp
```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 Use scraping function to extract and plot Cary data from 2018-2023

#set the years to scrape for
cary_years = rep(2018:2023)

# create a list of the same length as cary_years of the cary pwsid
cary_pwsid = rep('03-92-020', 
                 length(cary_years))


# use scraping function to pull data for all the years and bind_rows
df_caryyears <- map2(cary_pwsid, 
                     cary_years, 
                     scrape_func) %>% 
  bind_rows()

# plot Cary's max withdrawal by month for 2018 - 2023
plot_caryyears <- df_caryyears %>% 
  ggplot(aes(x=Date, y=Maximum_Day_Use))+
         #, color = as.factor(Year), group = Year))+
  geom_line(linewidth = 0.7)+
  geom_smooth(method="loess",se=FALSE, linewidth = 0.7)+
  scale_x_date(date_labels = paste0("%b"," '","%y"),
               date_breaks= "2 month")+
  labs(title = "Maximum Daily Withdrawal of Water",
       subtitle= paste("Cary, 2018 - 2023"),
       x = "Month and Year", 
       y = "Maximum Daily Use (MGD)")+
  theme(axis.text.x = element_text(angle = 70,
                                   hjust = 1, 
                                   size =8))
# scale_color_viridis_d()

plot_caryyears

```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: Cary does appear to have a trend in water usage over time. The trend line shows an increase in water usage from about 23MGD from 2018-2020 to about 24MGD from 2021-2023. 
>
